Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-01-29 doc/invt.tex
editing invt.tex

\begin{abstract}
  The Wang-Landau (WL) algorithm and metadynamics
  are two closely-related techniques in free energy simulations.
  %
  They allow one to sample a flat distribution
  along a quantity of interest, $z$,
  by adaptively constructing a bias potential
  that offsets the potential of mean force (PMF).
  %
  The two algorithms differ by
  the manner of updating the bias potential:
  %
  in the WL case, each simulation step triggers
  an update of the bias potential to only
  the bin containing the current $z$,
  while in the metadynamics case,
  the bias potential of several neighboring bins
  are affected as well.
  %
  The asymptotic error of the resulting PMF
  is determined by the updating magnitude,
  and the rate of the reduction of the magnitude
  over the simulation time.
  %
  In the WL case,
  a simple and effective formula for the magnitude
  is given by the inverse of the number of simulation steps.
  %
  In this study, we shall reexamine the optimality of this prescription,
  and attempt to extend it to the metadynamics case.
  %
  Further, we compared the two algorithms, and found that
  although metadynamics produces a smoother PMF profile,
  which is necessary for stable numerical differentiation
  in computing the force,
  it is less effective in reducing errors with the shorter wavelengths.
  %
  Thus, it relies more heavily on the assumption of
  the smoothness of the PMF profile,
  especially with a wide updating window.
\end{abstract}

@@

Free energy calculation\cite{frenkel} is a central theme
in computational physics and chemistry.
%
Given a system,
the problem is to compute a distribution, $p^*(z)$,
along a quantity of interest $z$, and to see how
the distribution changes with external conditions,
such as temperatures and pressures.
%
The negative logarithm $-\log p^*(z)$,
or the potential of mean force (PMF),
defines a free energy.
%
A straightforward approach of computing the PMF
is to run multiple independent simulations
under different conditions,
and then piece together the information
gathered there.
%
This solution is often unsatisfactory
for a complex and glassy system,
because the system can be trapped
in a local free energy minimum
for a long time.

A better solution is to artificially alter
the target distribution,
such that we can, in a single simulation, sample
a flat distribution\cite{mezei1987, berg1992, lee1993,
wang2001, wang2001pre, laio2002}
along $z$ over a wide range.
%
This alleviates the above problem of local trapping
as the system is now forced to travel more flexibly
along the $z$ direction,
which often represents a slow reaction coordinate.
%
It, however, requires one to efficiently construct a
bias potential for an often unknown system.
%
Note that the bias potential must exactly cancel the PMF
to achieve a flat distribution.

@@

\subsubsection{Integrals of the correlation functions}

@@

Another special case is the one-step process\cite{vankampen}:
in each step, the system hops to the previous bin $i - 1$
or the next bin $i + 1$ with equal probability $\frac 1 2$,
ignoring out-of-boundary transitions.
%
The transition matrix is
%
\begin{equation}
\arraycolsep=3.6pt\def\arraystretch{1.4}
\mathbf T
=
\left(
  \begin{array}{cccccccc}
    \frac 1 2 & \frac 1 2 & 0 & \dots & 0 \\
    \frac 1 2 & 0         & \frac 1 2 & \dots & 0 \\
    \vdots & &  & & \vdots \\
    0 & \dots & \frac 1 2 & 0  & \frac 1 2 \\
    0 & \dots & 0 & \frac 1 2 & \frac 1 2
  \end{array}
\right).
\label{eq:T_tridiag}
\end{equation}
%
We shall further assume that
the eigenvectors of $\mathbf w$
coincide with those of $\mathbf T$,
while the eigenvalues may differ
(this is true for a class of matrices
discussed in Sec. \ref{sec:band-matrix}).
%
The eigenvalues of $\mathbf T$ are
$$
\Lambda_k = 1 - 2 \sin^2 \frac{ (k - 1) \, \pi } { 2 \, n }.
$$
Note that these $\Lambda_k$'s are
not to be confused with the eigenvalues of $\mathbf w$,
$\lambda_k$'s.
%
The correlation function $\kappa_k(t)$
decays as $\kappa_k(0) \, \Lambda_k^t$,
with the same-time value $\kappa_k(0) = 1$ being the same as
that in perfect sampling.
%
Thus,
$\Gamma_k$ is roughly twice
the autocorrelation time,
and from Eq. \eqref{eq:Gamma_sum}, we get
%
\begin{equation}
\Gamma_k
=
1 + 2 \, \sum_{t = 1}^\infty \Lambda_k^t
=
\frac{ 1 + \Lambda_k } { 1 - \Lambda_k }
=
\cot^2 \left[ \frac{ (k - 1) \, \pi } { 2 \, n } \right].
\label{eq:Gamma_onestep}
\end{equation}
%
We can use this case to model local sampling processes.

@@

For an arbitrary updating matrix $\mathbf w$,
optimizing Eq. \eqref{eq:error_asym1}
can lead to an algebraically complex expression for $\alpha(t)$.
%
In practice, it is more useful to find a simple expression for $\alpha(t)$
that is near optimal.
%
From Eq. \eqref{eq:alpha_opt},
a reasonable candidate would be
%
\begin{equation}
\alpha(\tau) = \frac{1}{\lambda \, (\tau + t_0) },
\label{eq:alpha_trial}
\end{equation}
%
for some value of $\lambda$.
%
Below we shall assume this functional form
and find the optimal value
in some cases.
%
We shall show that the optimal $\lambda$
generally depends on not only the spectrum
of eigenvalues and the sampling method,
but also other parameters,
such as the simulation length, $t$.

Using Eq. \eqref{eq:alpha_trial}
in Eq. \eqref{eq:error_asym1} yields
%
\begin{align}
\Err_A
&=
\frac{    1    }
     { t + t_0 }
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k^2 }
       {    2 \, \nu_k - 1   }
\left[
  1 - \left(
        1 + \frac{ t }{ t_0 }
      \right)^{1 - 2 \, \nu_k}
\right],
\label{eq:error_asym_invt}
\end{align}
%
where $\nu_k \equiv \lambda_k / \lambda$.
%
At long times, we get
$$
\begin{aligned}
  \Err_A
  \stackrel{         t \to \infty         }
           { =\joinrel=\joinrel=\joinrel= }
  \sum_{k = 2}^n
  \begin{dcases}
    \frac{    1    }
         { t + t_0 }
    \frac{ \Gamma_k \, \nu_k^2 }
         {   2 \, \nu_k - 1    }
    &
    \mathrm{if \;} 2 \, \nu_k > 1,
    \\%[1em]
    %
    %
    \frac{    \Gamma_k    }
         { 4 \, (t + t_0) }
    \ln \frac{ t + t_0 }
             {   t_0  }
    &
    \mathrm{if \;} 2 \, \nu_k = 1,
    \\
    %
    %
    \frac{  t_0^{ 2 \, \nu_k  - 1}  }
         { (t + t_0)^{ 2 \, \nu_k } }
    \frac{ \Gamma_k \, \nu_k^2 }
         {   1 - 2 \, \nu_k    }
    &
    \mathrm{if \;} 2 \, \nu_k < 1.
  \end{dcases}
\end{aligned}
$$
%
The last two cases, $\lambda \ge 2\lambda_k$,
are asymptotically suboptimal
as they decay slower than $1/t$.
%
Thus, the optimal $\lambda$
satisfies $\lambda < 2 \min \lambda_k$,
and in this case,
we can seek that
the optimal value from
the smallest real root of
%
\begin{equation}
\sum_{k = 2}^n
\frac{ \lambda_k - \lambda }
{ \left(2 - \lambda/ \lambda_k \right)^2 }
= 0.
\label{eq:optimal_lambda_approx}
\end{equation}

@@

We give several remarks on Eq. \eqref{eq:error_invt}.
%
First, the final error is always greater than
the saturated equilibrium value at the terminal $\alpha(t)$:
$$
E > E_\mathrm{sat} \equiv \sum_{k = 2}^n E^f_k,
$$
because the second term in the braces
is nonnegative no matter the sign of $2 \, \nu_k - 1$
(in fact, it is a decreasing function
that vanishes only at $\nu_k \to +\infty$).

@@

Second, we shall show that the optimal $\lambda$
depends on the simulation length,
and it tends to decrease as simulation lengthens.
%
Let us first rewrite Eq. \eqref{eq:error_invt} as
%
$$
\Err
=
\sum_{ k = 2 }^n
  \frac{    \Gamma_k    }
       { 2 \, (t + t_0) }
  G\left( \nu_k, \frac{ t_0 } { t + t_0} \right),
$$
%
where the $k$th component is proportional to
%
$$
G\left( \nu_k, r \right)
=
\nu_k
\,
\left(
  1
  +
  \frac{ 1 - r^{2 \, \nu_k - 1} }
       {   2 \, \nu_k - 1       }
\right).
$$
%
As shown in Fig. \ref{fig:err_component},
the function $G(\nu_k, r)$ depends nontrivial on $\nu_k$.
%
For a fixed and sufficiently small $r = t_0 / (t + t_0)$,
$G(\nu_k, r)$ has two local minima along $\nu_k$,
one at $\nu_k = 0$, and the other around $\nu_k = 1$,
with the former being deeper albeit narrower.
%
This shows that for a relatively short simulation,
hence relative large value of $r$,
the minimal error may be achieved at $\nu_k \approx 0$,
corresponding to a large value of $\lambda$
($\lambda \gg \lambda_k$).
%
However, this minimum is unstable as simulation lengthens,
for $G(\nu_k, r)$ increases rapidly
with decreasing $r$ under a fixed $\nu_k$.
%
In other words,
asymptotically, the minimal error is always achieved at
$\nu_k = 1$ or $\lambda = \lambda_k$.
%
In this case,
$G(\nu_k, r) \to G(1, 0) = 2$
and the total error is roughly twice $E_\mathrm{sat}$.

@@

Particularly, if $\mu_2 = \dots = \mu_b = 0$,
the matrix $\mathbf w$ is tridiagonal,
%%
%\begin{equation}
%\arraycolsep=3.6pt\def\arraystretch{1.4}
%\mathbf w
%=
%\left(
%  \begin{array}{cccccccc}
%    1 - \mu_1   & \mu_1 & 0 & \dots & 0 \\
%    \mu_1 & 1 - 2 \, \mu_1  & \mu_1 & \dots & 0 \\
%    \vdots & &  & & \vdots \\
%    0 & \dots & \mu_1 & 1 - 2 \, \mu_1  & \mu_1 \\
%    0 & \dots & 0 & \mu_1 & 1 - \mu_1
%  \end{array}
%\right).
%\label{eq:wtridiag}
%\end{equation}
%%
%We have
and the eigenvalues are
\begin{equation}
  \lambda_k = 1 - 4 \, \mu_1 \sin^2 \frac{(k - 1) \, \pi}{2n}.
\label{eq:wtridiag_eigenvalue}
\end{equation}
%
Thus, for this updating matrix to be stable,
we need
$\min \lambda_k = \lambda_n > 0$,
and
$$
\mu_1 < \frac{1}{4 \, \cos^2\frac{\pi}{2n} }.
$$

We can find an analytical expression
of the optimal value of $\lambda$
in the limit of large $n$.
%
In this case,
we convert Eq.
\eqref{eq:error_asym_invt}
%and
%\eqref{eq:optimal_lambda_approx}
to an integral,
$\sum_i \to \frac{2 \, n}{\pi} \int dp$, and
$$
\begin{aligned}
\Err_2
&=
\frac{2 \, n}{\pi \, t}
\int_0^{\pi/2}
  \frac{    \Gamma(p) \, \left(1 - 4 \, \mu_1 \, \sin^2 p \right)^2    }
       { \lambda \, \left(2 - 8 \, \mu_1 \, \sin^2 p - \lambda \right) }
\, dp.
\end{aligned}
$$

For perfect sampling,
we have, from Eq. \eqref{eq:Gamma_perfect},
$\Gamma(p) = 1$, and
$$
\begin{aligned}
\Err_2
&=
\frac{2 \, n}{\pi \, t}
\int_0^{\pi/2}
\frac{ \left(1 - 4 \, \mu_1 \, \sin^2 p \right)^2 }
{ \lambda \, \left(2 - 8 \, \mu_1 \, \sin^2 p - \lambda \right) }
\, dp
\notag \\
&=
\frac{n}{4 \, t}
\left(
  \frac{2 - 4 \, \mu_1 + \lambda}{ \lambda }
  +
  \frac{ \lambda }
  { \sqrt{ (2 - \lambda) (2 - 8 \, \mu_1 - \lambda) } }
\right).
\end{aligned}
$$
\note{The integral is evaluated by contour integration.
%
It is somewhat more convenient to change variable $p \to \frac{ \pi } { 2 } - p$,
and
$$
\begin{aligned}
\Err_2
&=
\frac{n}{\lambda \, t}
\frac{1}{2 \pi i}
\int_0^{2 \, \pi}
\frac{ \left(1 - 4 \, \mu_1 \, \cos^2 p \right)^2 }
{ 2 - 8 \, \mu_1 \, \cos^2 p - \lambda }
\, dp
\\
&=
\frac{n}{\lambda \, t}
\frac{1}{2 \pi i}
\oint
\frac{ \left[1 - \mu_1 \, \left(z+\frac{1}{z}\right)^2 \right]^2 }
{ 2 - 2 \, \mu_1 \, \left(z + \frac{1}{z}\right)^2 - \lambda }
\, \frac{dz}{z},
\end{aligned}
$$
where the contour is the unit circle
in the complex plane of $z$.
%
This integral has five poles, one at the origin,
which produces a residue $\frac{2 - 4 \, \mu_1 - \lambda}{4}$
(from series expansion of small $z$),
and the other four at
$$
z + \frac{1}{z} = \pm\sqrt\frac{2-\lambda}{2 \, \mu_1},
$$
with two of them inside the unit circle:
$$
z_\pm = \pm \frac{\sqrt{2-\lambda} -\sqrt{2 - 8 \, \mu_1 - \lambda}}
{\sqrt{8 \mu_1}}.
$$
Thus,
$$
\begin{aligned}
\Err_2
&=
\frac{n}{\lambda \, t}
\left(
 \frac{2 - 4 \, \mu_1 - \lambda}{4}
 +
 \sum_{z = z_{\pm} }
 \frac{ \left(z^2 - \mu_1 (1 + z^2)^2 \right)^2 }
 { 2 z^4 (2 - 4 \mu_1 - \lambda - 4 \mu_1 z^2) }
\right)
\\
&=
\frac{n}{\lambda \, t}
\left(
  \frac{2 - 4 \, \mu_1 - \lambda}{4}
 +
 \frac{ \left(1 - \mu_1 \left(z + \frac{1}{z} \right)^2 \right)^2 }
 { 2 - 4 \mu_1 - \lambda - 4 \mu_1 z_{\pm}^2 }
\right)
\\
&=
\frac{n}{\lambda \, t}
\left(
  \frac{2 - 4 \, \mu_1 - \lambda}{4}
 +
 \frac{ (\lambda/2)^2 }
 { \sqrt{(2-\lambda) (2 - 8 \mu_1 -\lambda)} }
\right).
\end{aligned}
$$
}
%
Minimizing the function yields
%
\begin{equation}
\lambda = \frac{1 - 4 \, \mu_1} { 1 - 2 \, \mu_1 },
\label{eq:lambda_tridiag_perfect}
\end{equation}
%
and the error in the optimal is given by
%
\begin{equation}
\Err
=
\frac{n}{t}
\left(
  1+ \frac{2 \, \mu_1^2}{1-4 \, \mu_1}
\right).
\label{eq:error_tridiag_prefect}
\end{equation}
%
%From Eq. \eqref{eq:error_tridiag}, it is clear that
Clearly, the error
increases rapidly with magnitude of $\mu_1$,
no matter its sign,
and the minimal value, obtained at $\mu_1 = 0$,
corresponds to the single-bin scheme.

For one-step sampling,
we have, from Eq. \eqref{eq:Gamma_onestep},
$\Gamma(p) = \cot^2 p \approx \cos^2 p / (\sin^2 p + \delta^2)$,
where we have introduced
$\delta \propto \sin[ \pi / (2 \, n) ] \sim 1/n$
as a small parameter to avoid the divergence
around the origin.
%
Then
$$
\begin{aligned}
\Err_2
&=
\frac{   2 \, n }
     { \pi \, t }
\int_0^{\pi/2}
    \frac{            \left(1 - 4 \, \mu_1 \, \sin^2 p \right)^2         }
         { \lambda \, \left(2 - 8 \, \mu_1 \, \sin^2 p - \lambda \right) }
    \frac{ \cos^2 p }
         { \sin^2 p + \delta^2 }
\, dp
\notag \\
&
\stackrel{    \delta \ll 1     }
         { =\joinrel=\joinrel= }
\frac{   2 \, n }
     { \pi \, t }
\frac{             1             }
     {  \lambda \, (2 - \lambda) }
\frac{    1   }
     { \delta }
.
\end{aligned}
$$
%
The optimal $\lambda$ is therefore
%
\begin{equation}
\lambda = 1,
\label{eq:lambda_tridiag_onestep}
\end{equation}
%
which is the same of the global sampling scheme.

>>>>

\begin{abstract}
  The Wang-Landau (WL) algorithm and metadynamics
  are two closely-related techniques in free energy simulations.
  %
  They allow one to sample a flat distribution
  along a quantity of interest, $z$,
  by adaptively constructing a bias potential
  that offsets the potential of mean force (PMF).
  %
  The two algorithms differ by
  the manner of updating the bias potential:
  %
  in the WL case, each simulation step triggers
  an update of the bias potential to only
  the bin containing the current $z$,
  while in the metadynamics case,
  the bias potential of several neighboring bins
  are affected as well.
  %
  The asymptotic error of the resulting PMF
  is determined by the updating magnitude,
  and the rate of the reduction of the magnitude
  over the simulation time.
  %
  In the WL case,
  a simple and effective formula for the magnitude
  is given by the inverse of the number of simulation steps.
  %
  In this study, we shall reexamine the optimality of this prescription,
  and attempt to extend it to the metadynamics case.
  %
  Further, we compared the two algorithms, and found that
  although metadynamics produces a smoother PMF profile,
  it is less effective in reducing errors with the shorter wavelengths.
  %
  Thus, the latter relies more heavily on the assumption of
  the smoothness of the PMF profile,
  especially with a wide updating window.
\end{abstract}

@@

Free energy calculation\cite{frenkel} is a central theme
in computational physics and chemistry.
%
Given a system,
the problem is to compute,
via either Monte Carlo (MC) and
molecular dynamics (MD) simulations,
a distribution, $p^*(z)$,
along a quantity of interest $z$, and to see how
the distribution changes with external conditions,
such as temperatures and pressures.
%
The negative logarithm $-\log p^*(z)$,
or the potential of mean force (PMF),
defines a free energy.
%
A straightforward approach of computing the PMF
is to run multiple independent simulations
under different conditions,
and then piece together the information
gathered there.
%
This solution is often unsatisfactory
for a complex and glassy system,
because the system can be trapped
in a local free energy minimum
for a long time.

A better solution is to artificially alter
the target distribution,
such that we can, in a single simulation, sample
a flat distribution\cite{mezei1987, berg1992, lee1993,
wang2001, wang2001pre, laio2002, laio2008, barducci2011, sutto2012}
along $z$ over a wide range.
%
This alleviates the above problem of local trapping
as the system is now forced to travel more flexibly
along the $z$ direction,
which often represents a slow reaction coordinate.
%
It, however, requires one to efficiently construct a
bias potential for an often unknown system.
%
Note that the bias potential must exactly cancel the PMF
to achieve a flat distribution.

@@

\subsubsection{\label{sec:Gamma}
Integrals of the correlation functions}

@@

Another special case is the one-step process\cite{vankampen}:
in each step, the system hops to the previous bin $i - 1$
or the next bin $i + 1$ with equal probability $\frac 1 2$,
ignoring out-of-boundary transitions.
%
The transition matrix is
%
\begin{equation}
\arraycolsep=3.6pt\def\arraystretch{1.4}
\mathbf T
=
\left(
  \begin{array}{cccccccc}
    \frac 1 2 & \frac 1 2 & 0 & \dots & 0 \\
    \frac 1 2 & 0         & \frac 1 2 & \dots & 0 \\
    \vdots & &  & & \vdots \\
    0 & \dots & \frac 1 2 & 0  & \frac 1 2 \\
    0 & \dots & 0 & \frac 1 2 & \frac 1 2
  \end{array}
\right).
\label{eq:T_nn}
\end{equation}
%
We shall further assume that
the eigenvectors of $\mathbf w$
coincide with those of $\mathbf T$,
while the eigenvalues may differ
(this is true for a class of matrices
discussed in Sec. \ref{sec:band-matrix}).
%
The eigenvalues of $\mathbf T$ are
$$
\Lambda_k = 1 - 2 \sin^2 \frac{ (k - 1) \, \pi } { 2 \, n }.
$$
Note that these $\Lambda_k$'s are
not to be confused with the eigenvalues of $\mathbf w$,
$\lambda_k$'s.
%
The correlation function $\kappa_k(t)$
decays as $\kappa_k(0) \, \Lambda_k^t$,
with the same-time value $\kappa_k(0) = 1$ being the same as
that in perfect sampling.
%
Thus,
$\Gamma_k$ is roughly twice
the autocorrelation time,
and from Eq. \eqref{eq:Gamma_sum}, we get
%
\begin{equation}
\Gamma_k
=
1 + 2 \, \sum_{t = 1}^\infty \Lambda_k^t
=
\frac{ 1 + \Lambda_k } { 1 - \Lambda_k }
=
\cot^2 \left[ \frac{ (k - 1) \, \pi } { 2 \, n } \right].
\label{eq:Gamma_onestep}
\end{equation}
%
We can use this case to model local sampling processes.

@@

For an arbitrary updating matrix $\mathbf w$,
direct optimization of Eq. \eqref{eq:error_asym1}
can lead to an algebraically complex expression for $\alpha(t)$.
%
In practice, it is more useful to find
a simple expression for $\alpha(t)$
that is near optimal.
%
From Eq. \eqref{eq:alpha_opt},
a reasonable candidate would be
%
\begin{equation}
\alpha(\tau) = \frac{1}{\lambda \, (\tau + t_0) },
\label{eq:alpha_invtlambda}
\end{equation}
%
for some value of $\lambda$.
%
Below we shall assume this functional form
and find the optimal value
in some cases.
%
We shall show that the optimal $\lambda$
generally depends on not only the spectrum
of eigenvalues and the sampling method,
but also other parameters,
such as the simulation length, $t$.

Using Eq. \eqref{eq:alpha_invtlambda}
in Eq. \eqref{eq:error_asym1} yields
%
\begin{align}
\Err_A
&=
\frac{    1    }
     { t + t_0 }
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k^2 }
       {    2 \, \nu_k - 1   }
\left[
  1 - \left(
        1 + \frac{ t }{ t_0 }
      \right)^{1 - 2 \, \nu_k}
\right],
\label{eq:error_asym_invt}
\end{align}
%
where $\nu_k \equiv \lambda_k / \lambda$.
%
At long times, we get
$$
\begin{aligned}
  \Err_A
  \stackrel{         t \to \infty         }
           { =\joinrel=\joinrel=\joinrel= }
  \sum_{k = 2}^n
  \begin{dcases}
    \frac{    1    }
         { t + t_0 }
    \frac{ \Gamma_k \, \nu_k^2 }
         {   2 \, \nu_k - 1    }
    &
    \mathrm{if \;} 2 \, \nu_k > 1,
    \\%[1em]
    %
    %
    \frac{    \Gamma_k    }
         { 4 \, (t + t_0) }
    \ln \frac{ t + t_0 }
             {   t_0  }
    &
    \mathrm{if \;} 2 \, \nu_k = 1,
    \\
    %
    %
    \frac{  t_0^{ 2 \, \nu_k  - 1}  }
         { (t + t_0)^{ 2 \, \nu_k } }
    \frac{ \Gamma_k \, \nu_k^2 }
         {   1 - 2 \, \nu_k    }
    &
    \mathrm{if \;} 2 \, \nu_k < 1.
  \end{dcases}
\end{aligned}
$$
%
The last two cases, $\lambda \ge 2\lambda_k$,
are asymptotically suboptimal
as they decay slower than $1/t$.
%
Thus, the optimal $\lambda$
satisfies $\lambda < 2 \min \lambda_k$,
and in this case,
we can seek that
the optimal value from
the smallest real root of
%
\begin{equation}
\sum_{k = 2}^n
\frac{ \lambda_k - \lambda }
{ \left(2 - \lambda/ \lambda_k \right)^2 }
= 0.
\label{eq:optimal_lambda_approx}
\end{equation}

@@

%We give several remarks on Eq. \eqref{eq:error_invt}.
%

Clearly, the final error is always greater than
the saturated equilibrium value at the terminal $\alpha(t)$:
$$
E > E_\mathrm{sat} \equiv \sum_{k = 2}^n E^f_k,
$$
because the second term in the braces
is nonnegative no matter the sign of $2 \, \nu_k - 1$
(in fact, it is a decreasing function
that vanishes only at $\nu_k \to +\infty$).

\subsubsection{\label{sec:optlambda}
Optimal parameter $\lambda$}

We shall show that the optimal $\lambda$
depends on the simulation length,
and it tends to decrease as simulation lengthens.
%
Let us first rewrite Eq. \eqref{eq:error_invt} as
%
$$
\Err
=
\sum_{ k = 2 }^n
  \frac{    \Gamma_k    }
       { 2 \, (t + t_0) }
  G\left( \nu_k, \frac{ t_0 } { t + t_0} \right),
$$
%
where the $k$th component is proportional to
%
$$
G\left( \nu_k, r \right)
=
\nu_k
\,
\left(
  1
  +
  \frac{ 1 - r^{2 \, \nu_k - 1} }
       {   2 \, \nu_k - 1       }
\right).
$$
%
As shown in Fig. \ref{fig:err_component},
the function $G(\nu_k, r)$ depends non-trivially on $\nu_k$.
%
For a fixed and sufficiently small $r = t_0 / (t + t_0)$,
$G(\nu_k, r)$ has two local minima along $\nu_k$,
one at $\nu_k = 0$, and the other around $\nu_k = 1$,
with the former being deeper albeit narrower.
%
This means that for a relatively short simulation,
hence a relatively large value of $r$,
the minimal error may be achieved at $\nu_k \approx 0$,
corresponding to a large value of $\lambda$
($\lambda \gg \lambda_k$).
%
However, this minimum is unstable as simulation lengthens,
for $G(\nu_k, r)$ increases rapidly
with decreasing $r$ under a fixed $\nu_k$.
%
In other words,
asymptotically, the minimal error is always achieved at
$\nu_k = 1$ or $\lambda = \lambda_k$.
%
In this case,
$G(\nu_k, r) \to G(1, 0) = 2$
and the total error is roughly twice $E_\mathrm{sat}$.

For an arbitrary spectrum of eigenvalues,
$\lambda_2, \dots, \lambda_n$,
we expect that the optimal $\lambda$ to shift gradually to
smaller values as the simulation length, $t$, increasing.
%
Asymptotically, we expect the optimal $\lambda$
to be around $\min_k \lambda_k$.

@@

@@

Particularly, if $\mu_2 = \dots = \mu_b = 0$,
the matrix $\mathbf w$ is tridiagonal,
%%
%\begin{equation}
%\arraycolsep=3.6pt\def\arraystretch{1.4}
%\mathbf w
%=
%\left(
%  \begin{array}{cccccccc}
%    1 - \mu_1   & \mu_1 & 0 & \dots & 0 \\
%    \mu_1 & 1 - 2 \, \mu_1  & \mu_1 & \dots & 0 \\
%    \vdots & &  & & \vdots \\
%    0 & \dots & \mu_1 & 1 - 2 \, \mu_1  & \mu_1 \\
%    0 & \dots & 0 & \mu_1 & 1 - \mu_1
%  \end{array}
%\right).
%\label{eq:wnn}
%\end{equation}
%%
%We have
and the eigenvalues are
\begin{equation}
  \lambda_k = 1 - 4 \, \mu_1 \sin^2 \frac{(k - 1) \, \pi}{2n}.
\label{eq:wnn_eigenvalue}
\end{equation}
%
Thus, for this updating matrix to be stable,
we need
$\min \lambda_k = \lambda_n > 0$,
and
$$
\mu_1 < \frac{1}{4 \, \cos^2\frac{\pi}{2n} }.
$$

We can find an analytical expression
of the optimal value of $\lambda$
in the limit of large $n$.
%
In this case,
we convert Eq.
\eqref{eq:error_asym_invt}
%and
%\eqref{eq:optimal_lambda_approx}
to an integral,
$\sum_i \to \frac{2 \, n}{\pi} \int dp$, and
%
\begin{align}
\Err_A
=
\frac{2 \, n}{\pi \, t}
\int_0^{\pi/2}
  \frac{ \Gamma(p) \, \left(1 - 4 \, \mu_1 \, \sin^2 p \right)^2    }
       {   \lambda \, \left(2 - 8 \, \mu_1 \, \sin^2 p - \lambda \right) }
\, dp.
\notag
%\label{eq:error_nn_asym_int}
\end{align}

For perfect sampling,
we have, from Eq. \eqref{eq:Gamma_perfect},
$\Gamma(p) = 1$, and
$$
\begin{aligned}
\Err_A
&=
\frac{2 \, n}{\pi \, t}
\int_0^{\pi/2}
\frac{ \left(1 - 4 \, \mu_1 \, \sin^2 p \right)^2 }
{ \lambda \, \left(2 - 8 \, \mu_1 \, \sin^2 p - \lambda \right) }
\, dp
\notag \\
&=
\frac{n}{4 \, t}
\left(
  \frac{2 - 4 \, \mu_1 + \lambda}{ \lambda }
  +
  \frac{ \lambda }
  { \sqrt{ (2 - \lambda) (2 - 8 \, \mu_1 - \lambda) } }
\right).
\end{aligned}
$$
\note{The integral is evaluated by contour integration.
%
It is somewhat more convenient to change variable $p \to \frac{ \pi } { 2 } - p$,
and
$$
\begin{aligned}
\Err_A
&=
\frac{n}{\lambda \, t}
\frac{1}{2 \pi i}
\int_0^{2 \, \pi}
\frac{ \left(1 - 4 \, \mu_1 \, \cos^2 p \right)^2 }
{ 2 - 8 \, \mu_1 \, \cos^2 p - \lambda }
\, dp
\\
&=
\frac{n}{\lambda \, t}
\frac{1}{2 \pi i}
\oint
\frac{ \left[1 - \mu_1 \, \left(z+\frac{1}{z}\right)^2 \right]^2 }
{ 2 - 2 \, \mu_1 \, \left(z + \frac{1}{z}\right)^2 - \lambda }
\, \frac{dz}{z},
\end{aligned}
$$
where the contour is the unit circle
in the complex plane of $z$.
%
This integral has five poles, one at the origin,
which produces a residue $\frac{2 - 4 \, \mu_1 - \lambda}{4}$
(from series expansion of small $z$),
and the other four at
$$
z + \frac{1}{z} = \pm\sqrt\frac{2-\lambda}{2 \, \mu_1},
$$
with two of them inside the unit circle:
$$
z_\pm = \pm \frac{\sqrt{2-\lambda} -\sqrt{2 - 8 \, \mu_1 - \lambda}}
{\sqrt{8 \mu_1}}.
$$
Thus,
$$
\begin{aligned}
\Err_A
&=
\frac{n}{\lambda \, t}
\left(
 \frac{2 - 4 \, \mu_1 - \lambda}{4}
 +
 \sum_{z = z_{\pm} }
 \frac{ \left(z^2 - \mu_1 (1 + z^2)^2 \right)^2 }
 { 2 z^4 (2 - 4 \mu_1 - \lambda - 4 \mu_1 z^2) }
\right)
\\
&=
\frac{n}{\lambda \, t}
\left(
  \frac{2 - 4 \, \mu_1 - \lambda}{4}
 +
 \frac{ \left(1 - \mu_1 \left(z + \frac{1}{z} \right)^2 \right)^2 }
 { 2 - 4 \mu_1 - \lambda - 4 \mu_1 z_{\pm}^2 }
\right)
\\
&=
\frac{n}{\lambda \, t}
\left(
  \frac{2 - 4 \, \mu_1 - \lambda}{4}
 +
 \frac{ (\lambda/2)^2 }
 { \sqrt{(2-\lambda) (2 - 8 \mu_1 -\lambda)} }
\right).
\end{aligned}
$$
}
%
Minimizing the function yields
%
\begin{equation}
\lambda = \frac{1 - 4 \, \mu_1} { 1 - 2 \, \mu_1 },
\label{eq:lambda_nn_perfect}
\end{equation}
%
and the asymptotic error in the optimal case is given by
%
\begin{equation}
\Err_A
=
\frac{n}{t}
\left(
  1+ \frac{2 \, \mu_1^2}{1-4 \, \mu_1}
\right).
\label{eq:error_nn_prefect}
\end{equation}
%
%From Eq. \eqref{eq:error_nn}, it is clear that
Clearly, the error
increases with the magnitude of $\mu_1$
no matter its sign,
and the minimal value, obtained at $\mu_1 = 0$,
corresponds to the single-bin scheme.
%
Note that we have ignored the residual error
and assumed $\lambda < 2 \, \min \lambda_k = 2 - 8 \, \mu_1$
for all $k$ in the above solution.
%
Fortunately,
the $\lambda$ from Eq. \eqref{eq:lambda_nn_perfect}
satisfies this condition
by the stability condition,
$4 \, \mu_1 < 1$.

For the one-step sampling,
we find from Eq. \eqref{eq:Gamma_onestep}
that $\Gamma(p) = \cot^2 p$
is peaked around $p \approx 0$,
%
Then
$$
\begin{aligned}
\Err_A
&
\propto
\frac{   2 \, n }
     { \pi \, t }
\left.
\frac{            \left(1 - 4 \, \mu_1 \, \sin^2 p \right)^2         }
     { \lambda \, \left(2 - 8 \, \mu_1 \, \sin^2 p - \lambda \right) }
\right|_{ p \approx 0 }
\notag \\
&
\approx
\frac{   2 \, n }
     { \pi \, t }
\frac{             1             }
     {  \lambda \, (2 - \lambda) }
.
\end{aligned}
$$
%
The optimal $\lambda$ is therefore
%
\begin{equation}
\lambda \approx 1.
\label{eq:lambda_nn_onestep}
\end{equation}
%
